---
title: "Does Obama Veto Alot?"
output: html_document
---

Obama vetoed only 2 bills while in office, the least number of vetoes of any President in history. Is this proportionately a low number of vetoes? If uneventful congressional sessions had only produced 2 bills, than his veto rate would be 100%.

Let's look at the numbers.

```{r, echo = FALSE}
setwd('~/Projects/aimeerose.github.io/_drafts/bills')

# install.packages('RMySQL', type = 'source')
library(RMySQL)
library(ggplot2)
library(dplyr)
library(stringr)
library(reshape2)

conn <- dbConnect(MySQL(), user = 'root', password = '',
                  dbname = 'govtrack', host = 'localhost')

bills <- dbReadTable(conn, 'bills')

bills$congress <- factor(bills$congress)
bills$bill_type <- factor(bills$bill_type)

modern_bills <- bills[bills$congress %in% c(93:113), ]
```

## Has congress been busy?

A large majority of bills introduced into congress disappear into the legislative ether. At least 300,000 bills have been introduced into congress (govtrack.us). Each bill begins its journey in the introduced state, and progresses into as many as 29 different states. Most of these states signify ways a bill can die in Congress. For a complete listing of these statuses, see **Appendix A: 29 Flavors of Legislative State** at the bottom of this post.

In order to make these states digestable to the lay person, the states have been collapsed into:

* **Introduced:** The bill was introduced.
* **Referred:** The bill was referred to committee but never reported.
* **Reported:** The bill was addressed by its committee and reported to Congress.
* **Enacted:** The bill was enacted. This includes bills enacted by the ten-day rule and bills enacted by "unknown".
* **Enacted veto override:** The bill was enacted by Congressional override of a Presidential veto.
* **Passed bill:** The bill passed Congress but was never enacted before the end of the session.
* **Congressional stalemate:** The bill was never passed because of one of the following reasons: it did not receive mutual approval in the House and Senate; non-identical forms were passed; it was killed by a suspension vote; it failed vote for cloture or it was killed while resolving differences.
* **Vetoed:** The bill was vetoed successfully. This includes pocket vetoes.
* **Resolution passed:** Passed simple and concurrent resolutions.

### 69% of bills in US history remain in a "referred" state 

69% of bills introduced into congress are listed as in the "referred" state, waiting to be addressed by their committee. If only counting back to 1973, an even greater proportion, 81% of bills, are in the referred state.

```{r, echo = FALSE}
enacted <- list(
  'enacted_signed',
  'enacted_unknown',
  'enacted_tendayrule')

congressional_stalemates <- list(
  'pass_over_senate',
  'pass_over_house',
  'pass_back_house',
  'pass_back_senate',
  'fail_originating_house',
  'fail_originating_senate',
  'conference_passed_house',
  'fail_second_house',
  'fail_second_senate',
  'conference_passed_senate',
  'prov_kill_suspensionfailed',
  'prov_kill_cloturefailed')

vetoed <- list(
  'prov_kill_veto',
  'vetoed_pocket',
  'override_pass_over_house',
  'vetoed_override_fail_second_senate',
  'vetoed_override_fail_originating_house',
  'vetoed_override_fail_originating_senate',
  'vetoed_override_fail_second_house')

resolution_passed <- list(
  'passed_concurrentres',
  'passed_simpleres')

collapsed_states <- function(bill_state){
  returned_state <- if (bill_state %in% congressional_stalemates) {
    return('congressional_stalemates')
  } else if (bill_state %in% enacted) {
    return('enacted')
  } else if (bill_state %in% vetoed) {
    return('vetoed')
  } else if (bill_state %in% resolution_passed) {
    return('resolution_passed')
  } else {
    return(bill_state)
  }
  
  return(returned_state)
}

modern_bills$current_status.f <- unlist(lapply(modern_bills[,c('current_status')], collapsed_states))

grouped_bills <- modern_bills %>%
  group_by(current_status.f) %>%
  summarise(total = n()) %>%
  mutate(freq = round(total / sum(total), 4)) %>%
  arrange(total)

print(grouped_bills)

grouped_bills_subset <- subset(grouped_bills, total > 100)

ggplot(data = grouped_bills, aes(x = 1, y = total, fill = current_status.f)) +
  geom_bar(stat = 'identity', color = 'black') +
  coord_polar(theta='y', start = 45) +
  guides(fill = guide_legend(override.aes = list(colour = NA), title = 'Current Status')) +
  theme(axis.ticks = element_blank(), axis.title = element_blank(), axis.text.y = element_blank()) +
  theme(axis.text.x = element_blank()) +
  ggtitle('Current Bill Statuses since 93rd Congress')
```


### Claims congress's productivity is slowing are weakly supported by the number of bills enacted.

How effectual have the different congressional sessions been?

The referred state overwhelms visual analysis of the states by congress.

```{r, echo = FALSE}
ggplot(data = modern_bills, aes(x = congress, fill = current_status.f)) + 
  geom_bar() +
  ggtitle('Current State of Bills Introduced by Congressional Session') +
  scale_y_continuous('Total number of bills introduced', breaks = seq(0,30000,5000)) +
  labs(x = 'Congressional session') +
  guides(fill = guide_legend(override.aes = list(colour = NA), title = 'Current Status'))
```

Claims congress has become less effectual are weakly supported by the number of bills enacted.

```{r, echo = FALSE}
ggplot(data = subset(modern_bills, current_status.f %in% list('enacted', 'enacted_veto_override')), aes(x = congress)) + 
  geom_histogram(colour = 'darkolivegreen', fill = 'darkolivegreen3') +
  scale_y_continuous(breaks = seq(0, 800, 100)) +
  ggtitle('Number of bills enacted by Congress')
```

### Reagans congressional sessions' introduction of resolutions dwarves those before and after

What if productivity is not only number of bills enacted, but also the number of bills introduced?

```{r, echo = FALSE}
ggplot(data = subset(modern_bills, current_status.f == 'introduced'), aes(x = congress)) + 
  geom_histogram(colour = 'darkolivegreen', fill = 'darkolivegreen3') +
  ggtitle('Number of bills introduced to Congress')
```

Whoa! What was going on in the 97 - 100th Congressional sessions?

Congressional activity, specifically new resolutions, spiked during Ronald Reagan's congressional sessions. Comparing 4 sessions before, after and during the Reagan presidency revealed **nearly 10 times more house resolutions from the Reagan years still exist in the introduced state as before and after the Reagan years.** From those sessions, 996 house resolutions are still in the "introduced" state, whereas before and after Reagan sessions only 103 and 26 house resolutions still exist as "introduced", respectively. Bill sponsors mirror the imbalance: during the Reagan years, the top 10 sponsors of bills have at least 40 and as many as 125 introduced bills or resolutions. The top 10 before and after the Reagan sessions all have 17 or less outstanding.


```{r, echo = FALSE}
extract_sortname <- function(sponsor_blob) {
 extracted <- str_extract(sponsor_blob, "sortname\\\"(.+)\\\",")
 return(sub("\\\",", "", sub("sortname\\\"=>\\\"", "", extracted)))
}

modern_bills$sponsor.sort_name <- unlist(lapply(modern_bills[,c('sponsor')], extract_sortname))

modern_bills$sponsor.sort_name <- factor(modern_bills$sponsor.sort_name)

reagan_bills <- modern_bills[modern_bills$congress %in% c(97:100), ]
reagan_bills_introduced <- subset(reagan_bills, current_status == 'introduced')

# 170 levels
grouped_reagan_bills <- reagan_bills_introduced %>%
  group_by(sponsor.sort_name) %>%
  summarise(total = n()) %>%
  mutate(freq = round(total / sum(total), 4)) %>%
  arrange(-total)

before_reagan_bills <- modern_bills[modern_bills$congress %in% c(93:96), ]
before_reagan_bills_introduced <- subset(before_reagan_bills, current_status == 'introduced')

grouped_before_reagan_bills <- before_reagan_bills_introduced %>%
  group_by(sponsor.sort_name) %>%
  summarise(total = n()) %>%
  mutate(freq = round(total / sum(total), 4)) %>%
  arrange(-total)

after_reagan_bills <- modern_bills[modern_bills$congress %in% c(101:104), ]
after_reagan_bills_introduced <- subset(after_reagan_bills, current_status == 'introduced')

grouped_after_reagan_bills <- after_reagan_bills_introduced %>%
  group_by(sponsor.sort_name) %>%
  summarise(total = n()) %>%
  mutate(freq = round(total / sum(total), 4)) %>%
  arrange(-total)

# Before
grouped_before_reagan_bills <- before_reagan_bills_introduced %>%
  group_by(bill_type) %>%
  summarise(total = n()) %>%
  mutate(freq = round(total / sum(total), 4)) %>%
  arrange(-total)

pie_grouped_before_reagan <- ggplot(data = grouped_before_reagan_bills, aes(x = 1, y = total, fill = bill_type)) +
  geom_bar(stat = 'identity', color = 'black') +
  ggtitle('Introduced before Reagan') +
  coord_polar(theta='y', start = 45) +
  guides(fill = guide_legend(override.aes = list(colour = NA), title = 'Bill Types')) +
  theme(axis.ticks = element_blank(), axis.title = element_blank(), axis.text.y = element_blank()) +
  theme(axis.text.x = element_blank())

# After
grouped_after_reagan_bills <- after_reagan_bills_introduced %>%
  group_by(bill_type) %>%
  summarise(total = n()) %>%
  mutate(freq = round(total / sum(total), 4)) %>%
  arrange(-total)

pie_grouped_after_reagan <- ggplot(data = grouped_after_reagan_bills, aes(x = 1, y = total, fill = bill_type)) +
  geom_bar(stat = 'identity', color = 'black') +
  ggtitle('Introduced after Reagan') +
  coord_polar(theta='y', start = 45) +
  guides(fill = guide_legend(override.aes = list(colour = NA), title = 'Bill Types')) +
  theme(axis.ticks = element_blank(), axis.title = element_blank(), axis.text.y = element_blank()) +
  theme(axis.text.x = element_blank())

# During
grouped_reagan_bills <- reagan_bills_introduced %>%
  group_by(bill_type) %>%
  summarise(total = n()) %>%
  mutate(freq = round(total / sum(total), 4)) %>%
  arrange(-total)

pie_grouped_during_reagan <- ggplot(data = grouped_reagan_bills, aes(x = 1, y = total, fill = bill_type)) +
  geom_bar(stat = 'identity', color = 'black') +
  ggtitle('Introduced during Reagan') +
  coord_polar(theta='y', start = 45) +
  guides(fill = guide_legend(override.aes = list(colour = NA), title = 'Bill Types')) +
  theme(axis.ticks = element_blank(), axis.title = element_blank(), axis.text.y = element_blank()) +
  theme(axis.text.x = element_blank())

pie_grouped_before_reagan
pie_grouped_during_reagan
pie_grouped_after_reagan
```

## Does Obama veto a lot?

Obama was president during the 111th, 112th and 113th Congressional sessions.

```{r, echo = FALSE}
modern_bills_grouped <- modern_bills %>%
  group_by(congress, current_status.f) %>%
  summarise(total_of_status = n()) %>%
  mutate(total_bills = sum(total_of_status))

casted <- dcast(modern_bills_grouped, value.var = 'total_of_status', total_bills + congress ~ current_status.f)
casted$proportion_vetoed <- casted$vetoed / casted$total_bills

ggplot(data = casted, aes(x = congress, y = proportion_vetoed)) + 
  geom_bar(stat = 'identity', colour = 'darkolivegreen', fill = 'darkolivegreen3') +
  ylab('Proportion Vetoed') +
  ggtitle('Proportion of bills vetoed')
```

**Obama has issued, both in total numbers and relative to total bills, the least number of vetoes when compared with other Presidents.**

**There is also weak evidence to suggest the number of bills enacted, and thus bills available to veto, is declining over time.**

```{r, echo = FALSE}
ggplot(data = subset(modern_bills, current_status.f %in% list('enacted', 'enacted_veto_override')), aes(x = congress)) + 
  geom_histogram(colour = 'darkolivegreen', fill = 'darkolivegreen3') +
  scale_y_continuous(breaks = seq(0, 800, 100)) +
  ylab('Number of bills enacted') +
  ggtitle('Number of bills enacted by Congress')
```


### Appendix A: 29 Flavors of Legislative State

#### Identity
* **referred:** The bill was introduced and referred to a committee, but never reported.
* **reported:** The bill was adressed in a hearing by its committee, but was never enacted by Congress.
* **introduced:** The bill was introduced.
* **enacted_veto_override:** The bill was enacted after a congressional override of the Presidential veto.
* **passed_bill:** The bill passed Congress but was never enacted before the end of the session.

#### Congressional Stalemate
* **pass_over_senate:** The bill passed the Senate but not the House.
* **pass_over_house:** The bill passed the House but not Senate.
* **pass_back_house:** The bill passed by both chambers but was passed in non-identical forms and the differences were never resolved.
* **pass_back_senate:** The bill passed by both chambers but was passed in non-identical forms and the differences were never resolved.
* **fail_originating_house:** The bill failed in the House.
* **fail_originating_senate:** The bill failed in the Senate.
* **conference_passed_house:** The bill was introduced in a previous session of Congress. It was passed by both chambers but it passed in non-identical forms.
* **fail_second_house:** The bill passed in senate, failed in house.
* **failed_second_senate:** The bill was passed in the House but failed in the Sentate.
* **conference_passed_senate:** The bill was introduced in a previous session of Congress. It was passed in non-identical form and only one chamber approved a conference report to resolve the differences.
* **prov_kill_suspensionfailed:** A catch-all for random ways for a bill to die in Congress. The bill was introduced but was killed due to a failed vote for cloture (placing a time limit on the debate), under a fast-track vote called 'suspension' (meant to quickly pass non-controversial bills), or while resolving differences.
* **prov_kill_cloturefailed:** Same as description above for 'prov_kill_suspensionfailed'.

#### Vetoed
* **prov_kill_veto:** The bill was vetoed by the President and Congress did not attempt an override.
* **vetoed_pocket:** The bill was pocket vetoed. A bill is pocket vetoed when the President does not sign the bill and Congress adjourns before the ten-day period expires.
* **override_pass_over_house:** The bill was vetoed by the President and Congress did not finish an override before the end of the Congressional session.
* **vetoed_override_fail_second_senate:** The bill was vetoed. The Senate attempted to override the veto but failed.
* **vetoed_override_fail_originating_house:** The bill was vetoed. The House attempted to override the veto but failed.
* **vetoed_override_fail_originating_senate:** The bill was vetoed by the President and Senate did not finish an override before the end of the Congressional session.
* **vetoed_override_fail_second_house:** The bill was vetoed by the President and the House did not finish an override before the end of the Congressional session.

#### Enacted
* **enacted_signed:** The bill was enacted after being signed.
* **enacted_unknown:** The bill was enacted.
* **enacted_tendayrule:** The bill became enacted after ten days elapsed after being presented to the President.

#### Passed Resolutions
* **passed_concurrentres:** The concurrent resolution was agreed to by both Chambers, which is the end of the legislative process for concurrent resolutions.
* **passed_simpleres:** The bill is a simple resolution and was passed^.

***simple resolution:*** Designated "S. Res.," simple resolutions are used to express nonbinding positions of the Senate or to deal with the Senate's internal affairs, such as the creation of a special committee. They do not require action by the House of Representatives. ([source](https://www.senate.gov/reference/glossary_term/simple_resolution.htm))


### Resources

http://mathematicalcoffee.blogspot.com/2014/06/ggpie-pie-graphs-in-ggplot2.html